---
title: "Basic Exploratory Analysis of BEAM Outbreak Antimicrobial Isolate Data"
date: "2026-04-09"
output: html_document
---

# Setup

Load needed packages, make sure they are installed.

```{r}
library(readr) 
library(dplyr) 
library(tidyverse) 
library(skimr) 
library(here)
library(forecast)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

So for now in this qmd we are calling the preliminarily processed dataframe with 2118 rows and 13 columns. 

# Data exploration through figures

So I think first I just want to see plots, one for each year, with number of outbreak isolates per state, just to get a baseline. 

```{r}
##Grouping first 
year_state_subset <- mydata %>% group_by(Year = year(date), State) %>% summarise(isolate_count = sum(`Outbreak associated isolates`, na.rm = TRUE), .groups = "drop")
##Making faceted bar plots
plot1 <- ggplot(year_state_subset, aes(x = State, y = isolate_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~Year) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size = 4)) + 
  labs(x = "State", y = "Number of Outbreak Associated Isolates", title = "Outbreak Associated Isolates per State")

##Saving
figure_file = here("results", "figures", "faceted-states-isolates.png")
ggsave(filename = figure_file, plot = plot1) 

top_states <- year_state_subset %>% group_by(Year) %>% slice_max(order_by = isolate_count, n = 1, with_ties = FALSE) %>% select(State, isolate_count, Year)
print(top_states)

```

Then I want to see total number of outbreak isolates over time. 

```{r}
##To see the continuous plot over time that I want to see, I need to make a proper date column
df_monthly <- mydata %>%
  group_by(date) %>%
  summarise(total_isolates = sum(`Outbreak associated isolates`)) 
###Now making the plot
plot2 <- ggplot(df_monthly, aes(x = date, y = total_isolates)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months")+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Date", y = "Number of Outbreak Associated Isolates", title = "Outbreak Associated Isolates Over Time") 
##Saving
figure_file = here("results", "figures", "outbreaks-over-time.png")
ggsave(filename = figure_file, plot = plot2) 
```

I also want to see a genus/species breakdown of number of isolates, to see which bacterium has the most outbreaks.

```{r}
##To see the unique genus and species (Pathogen and Serotype/Species) combos, I need to make a new column first
df_naming <- mydata %>% group_by(genus_species) %>% summarise(total_outbreak_isolates = sum(`Outbreak associated isolates`, na.rm = TRUE)) %>% filter(total_outbreak_isolates > 0)
###Now making plot
plot3 <- ggplot(df_naming, aes(x = genus_species, y = total_outbreak_isolates)) +
  geom_col() +
  coord_flip() + 
  labs(x = "Pathogen", y = "Number of Outbreak Isolates", title = "Number of Outbreak Isolates per Pathogen") +
  theme(axis.text.x = element_text(size = 5)) +
  theme(axis.text.y = element_text(size = 2.5))

##Saving 
figure_file = here("results", "figures", "naming-number-isolates.png")
ggsave(filename = figure_file, plot = plot3) 

top_species <- df_naming %>% slice_max(order_by = total_outbreak_isolates, n = 20, with_ties = FALSE) %>% select(genus_species, total_outbreak_isolates)
print(top_species)

```

Also interested to see if source site or source type are in anyway correlated to number of outbreak isolates. 

```{r}
####I did these plots initially, before I did my last round of cleaning up my data frame, so these columns aren't included in the most recent version anymore.
###Straightforward plots this time 
#plot4 <- ggplot(mydata, aes(x = `Source Site`, y = `Outbreak associated isolates`)) + 
    #geom_col()

#plot5 <- ggplot(mydata, aes(x = `Source Type`, y = `Outbreak associated isolates`)) + 
    #geom_col()
#Saving
#figure_file = here("results", "figures", "site-outbreaks.png")
#ggsave(filename = figure_file, plot = plot4) 

#figure_file = here("results", "figures", "type-outbreaks.png")
#ggsave(filename = figure_file, plot = plot5) 
```


```{r}
###Need to investigate is antimicrobial resistance increasing over time, and which species are associated - this can also be looked at with just ggplot I think 
df_anti_per_iso_time <- mydata %>% distinct(genus_species, .keep_all = TRUE) %>% arrange(Year = year(date)) %>% mutate(`Percent Outbreak Isolates with AMR` = (`Number of Outbreak Isolates with antimicrobial resistance` / `Outbreak associated isolates`) * 100)

df_anti_per_iso_time <- df_anti_per_iso_time %>% drop_na(`Percent Outbreak Isolates with AMR`)
  
plot6 <- ggplot(df_anti_per_iso_time, aes(x = date, y = `Percent Outbreak Isolates with AMR`, color = genus_species, group = genus_species)) +
  geom_line(size = 1) + 
  geom_point() +
  labs(title = "Antimicrobial Resistance Over Time", x = "Date", y = "Percent of Outbreak Isolates with Antimicrobial Resistance", color = "Species") + scale_color_discrete(na.translate = FALSE) + ylim(0, 2)

figure_file = here("results", "figures", "resistance_time_species.png")
ggsave(filename = figure_file, plot = plot6) 

result <- df_anti_per_iso_time %>% slice_max(order_by = `Percent Outbreak Isolates with AMR`, n = 5, with_ties = FALSE) %>% select(genus_species, `Percent Outbreak Isolates with AMR`, date)
print(result)

df_anti_time <- mydata %>% group_by(date) %>% summarise(`Number Anti` = sum(`Number of Outbreak Isolates with antimicrobial resistance`, na.rm = TRUE)) %>%
  ungroup()

plot7 <- ggplot(df_anti_time, aes(x = date, y = `Number Anti`)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point()

figure_file = here("results", "figures", "resistnce_over_time.png")
ggsave(filename = figure_file, plot = plot7) 

current_state <- df_anti_time %>%
  summarise(
    current_value = last(`Number Anti`),
    overall_mean = mean(`Number Anti`),
    recent_mean = mean(tail(`Number Anti`, 6)),
    trend_slope = coef(lm(`Number Anti` ~ as.numeric(date)))[2],
    percent_change_from_start = (last(`Number Anti`) - first(`Number Anti`)) / first(`Number Anti`) * 100,
    is_increasing = last(`Number Anti`) > mean(`Number Anti`))

print(current_state)

forecast_model <- auto.arima(df_anti_time$`Number Anti`)
future_forecast <- forecast(forecast_model, h = 6)
last_date <- max(df_anti_time$date)
forecast_dates <- seq(last_date, by = "month", length.out = 7)[-1]

forecast_df <- data.frame(date = forecast_dates, forecast = as.numeric(future_forecast$mean), lower_80 = as.numeric(future_forecast$lower[,1]), upper_80 = as.numeric(future_forecast$upper[,1]), lower_95 = as.numeric(future_forecast$lower[,2]), upper_95 = as.numeric(future_forecast$upper[,2]))

# Plot with actual dates
plot8 <- ggplot() +
  geom_line(data = df_anti_time, aes(x = date, y = `Number Anti`), color = "black", linewidth = 1) +
  geom_point(data = df_anti_time, aes(x = date, y = `Number Anti`)) +
  geom_line(data = forecast_df, aes(x = date, y = forecast), color = "blue", linewidth = 1, linetype = "dashed") +
  geom_ribbon(data = forecast_df, aes(x = date, ymin = lower_95, ymax = upper_95), fill = "blue", alpha = 0.2) +
  geom_ribbon(data = forecast_df, aes(x = date, ymin = lower_80, ymax = upper_80), fill = "blue", alpha = 0.3) +
  theme_minimal() +
  labs(title = "Antimicrobial Resistance: Historical & Forecast", x = "Date", y = "Number of Resistant Isolates", caption = "Shaded areas represent 80% and 95% confidence intervals")

figure_file = here("results", "figures", "historical_forecast-resistance.png")
ggsave(filename = figure_file, plot = plot8) 



```