---
title: "Basic Exploratory Analysis of BEAM Outbreak Antimicrobial Isolate Data"
date: "2026-04-09"
output: html_document
---

# Setup

Load needed packages, make sure they are installed.

```{r}
library(readr) 
library(dplyr) 
library(tidyverse) 
library(skimr) 
library(here)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

So for now in this qmd we are calling the preliminarily processed dataframe with 2118 rows and 13 columns. 

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Data exploration through figures

So I think first I just want to see plots, one for each year, with number of outbreak isolates per state, just to get a baseline. 

```{r}
##Grouping first 
year_state_subset <- mydata %>% group_by(Year, State) %>%
    summarise(isolate_count = n(), .groups = 'drop')
##Making faceted bar plots
plot1 <- ggplot(year_state_subset, aes(x = State, y = isolate_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  facet_wrap(~Year) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1, size = 4))
##Saving
figure_file = here("results", "figures", "faceted-states-isolates.png")
ggsave(filename = figure_file, plot = plot1) 
```

Then I want to see total number of outbreak isolates over time. 

```{r}
##To see the continuous plot over time that I want to see, I need to make a proper date column
df_monthly <- mydata %>%
  mutate(date = make_date(Year, Month, 1)) %>%
  group_by(date) %>%
  summarise(total_isolates = sum(`Outbreak associated isolates`)) 
###Now making the plot
plot2 <- ggplot(df_monthly, aes(x = date, y = total_isolates)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months")+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
##Saving
figure_file = here("results", "figures", "outbreaks-over-time.png")
ggsave(filename = figure_file, plot = plot2) 
```

I also want to see a genus/species breakdown of number of isolates, to see which bacterium has the most outbreaks.

```{r}
##To see the unique genus and species (Pathogen and Serotype/Species) combos, I need to make a new column first
df_naming <- mydata %>%
  mutate(genus_species = paste(Pathogen, `Serotype/Species`, sep = " ")) %>%
  count(genus_species)
###Now making plot
plot3 <- ggplot(df_naming, aes(x = genus_species, y = n)) +
  geom_col() +
  coord_flip()
##Saving 
figure_file = here("results", "figures", "naming-number-isolates.png")
ggsave(filename = figure_file, plot = plot3) 
```

Also interested to see if source site or source type are in anyway correlated to number of outbreak isolates. 

```{r}
###Straightforward plots this time 
plot4 <- ggplot(mydata, aes(x = `Source Site`, y = `Outbreak associated isolates`)) + 
    geom_col()

plot5 <- ggplot(mydata, aes(x = `Source Type`, y = `Outbreak associated isolates`)) + 
    geom_col()
#Saving
figure_file = here("results", "figures", "site-outbreaks.png")
ggsave(filename = figure_file, plot = plot4) 

figure_file = here("results", "figures", "type-outbreaks.png")
ggsave(filename = figure_file, plot = plot5) 
```


```{r}
###Need to investigate is antimicrobial resistance increasing over time, and which species are associated - this can also be looked at with just ggplot I think 
df_anti_per_iso_time <- mydata %>% mutate(`Proportion of Outbreaks Isolates with Antimicroobial Resistance` = `Number of Outbreak Isolates with antimicrobial resistance`/`Outbreak associated isolates`) %>% 
  mutate(genus_species = paste(Pathogen, `Serotype/Species`, sep = " ")) %>% 
  mutate(date = make_date(Year, Month, 1)) %>%
  group_by(date)
  
plot6 <- ggplot(df_anti_per_iso_time, aes(x = Year, y = `Proportion of Outbreaks Isolates with Antimicroobial Resistance`)) +
  geom_bar(stat = "identity", fill = "steelblue") + 
  facet_wrap(~genus_species) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

figure_file = here("results", "figures", "faceted_resistance_time_species.png")
ggsave(filename = figure_file, plot = plot6) 

df_anti_time <- mydata %>% group_by(Year) %>% mutate(Number_Anti = sum(`Number of Outbreak Isolates with antimicrobial resistance`))

plot7 <- ggplot(df_anti_time, aes(x = Year, y = `Number_Anti`)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point()

figure_file = here("results", "figures", "resistnce_over_time.png")
ggsave(filename = figure_file, plot = plot7) 
```