---
title: "Cleaning Script"
author: "Elle Adams"
date: "2026-04-04"
output: html_document
---

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readr) 
library(dplyr) 
library(tidyverse) 
library(skimr) 
library(here)
```

# Data loading

```{r}
#Read in the raw data
data_location <- here("data","raw-data","BEAM_Dashboard_-_Report_Data_20260304.csv")
rawdata <- read_csv(data_location)
```

# Check data

```{r}
glimpse(rawdata)
summary(rawdata)
head(rawdata)
skim(rawdata)
```

Some basic information: 239,804 rows and 13 columns. 
Some columns have similar names, according to the CDC Beam website Number of multistate outbreaks = Number of new multistate outbreaks associated with each state and Number of multistate outbreaks - US = Number of unique new multistate outbreaks across the US.

# Cleaning

It appears that 7 of the rows have missing state data, I need to see which those are. 

```{r}
rows_missing_state <- which(is.na(rawdata$State))

print(rawdata[rows_missing_state, ])
```

Now I can see that rows 44453, 61088, 61089, 64372, 79239, 79240, and 80401 have missing State information.

The last five columns of data have a LOT of NAs, which I need to keep in mind for the analysis, but all of those columns should contain numbers, so I might be able to turn all of the NAs in zeros if that would help with analysis. First I kind of just want to see which rows actually have data in those columns.

```{r}
##Outbreak associated isolates column - identify which rows actually have data 
oai_rows_data <- rawdata %>%
  filter(complete.cases(rawdata$`Outbreak associated isolates`))

print(oai_rows_data) #20,208 rows actually have data

##New multistate outbreaks column 
nmo_rows_data <- rawdata %>%
  filter(complete.cases(rawdata$`New multistate outbreaks`))

print(nmo_rows_data) #9,922 rows actually have data

##New multistate outbreak - US column
nmou_rows_data <- rawdata %>%
  filter(complete.cases(rawdata$`New multistate outbreaks - US`))

print(nmou_rows_data) #1,266 rows actually have data

##% Isolates with clinically important antimicrobial resistance column
pcia_rows_data <- rawdata %>%
  filter(complete.cases(rawdata$`% Isolates with clinically important antimicrobial resistance`))

print(pcia_rows_data) #2,118 rows actually have data

##Number of sequenced isolates analyzed by NARMS column 
nsia_rows_data <- rawdata %>%
  filter(complete.cases(rawdata$`Number of sequenced isolates analyzed by NARMS`))

print(nsia_rows_data) #7,712 rows actually have data

##Number of rows that are complete
complete_rows_data <- rawdata %>%
    filter(complete.cases(rawdata))

print(complete_rows_data) # 147 rows are COMPLETE
```

Based on looking at this, it looks like really no matter what questions I decide to ask, I will have thousands of samples to contribute to the answer. 

```{r}

```

All done, data is clean now. 

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

```{r}
processeddata <- 
```


# Save data 

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```


